Estimation of flow trajectories in a multiple lines network

### 2022-04-26 presentation_bb/presentation_bb.tex
Presentation in pre-final form

\title{Estimation of flow trajectories in a multiple lines network}
\subtitle{Case studies with \textit{transports publics de la région lausannoise} (tl) data}
\date{}
\author{Guillaume Guex \\ Romain Loup \\ François Bavaud}
\institute{University of Lausanne}

@@

@@

@@

	\begin{frame}{Formal problem definition}
		Let a line (in one direction), which have $n$ stops, indexed by line order. Let $\bm{\rho}_\text{in} = (\rho^\text{in}_s)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_t)$ be two vectors representing, respectively, the \imp{passengers entering and leaving lines at each stop}. \\
		\vspace{0.4cm}
		We search a $(n \times n)$ \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$ where components represents
		$$
			n_{st} = \text{"\imp{the number of passengers entering line at $s$ and leaving at $t$}"}.
		$$
		These components must verify
		\begin{enumerate}
			\item $n_{st} \geq 0$,
			\item $n_{s\bullet} = \rho^\text{in}_s$,
			\item $n_{\bullet t} = \rho^\text{out}_t$.
		\end{enumerate}
		\small ($\bullet$ indicates a sum on the replaced index)
	\end{frame}

@@

	The same solution can be obtained with the \imp{iterative proportional fitting (IPF)} algorithm. Let 
	\begin{enumerate}
		\item $\mathbf{P} = (p_{ij})$ a $(n \times m)$ matrix, 
		\item $\mathbf{u} = (u_i)$ a n-length vector, and 
		\item $\mathbf{v} = (v_i)$ a m-length vector,
	\end{enumerate}
	all of them with strictly positive components. We can find two vectors $\mathbf{a} = (a_i)$ and $\mathbf{b} = (b_i)$ such that the matrix $\mathbf{Q} = (q_{ij})$, defined with
	$$
		q_{ij} = a_i b_j p_{ij},
	$$
	verifies
	\begin{itemize}
		\item $q_{i\bullet} = u_i$,
		\item $q_{\bullet j} = v_j$,
		\item $K(\mathbf{Q}| \mathbf{P}) := \sum_{ij} \frac{q_{ij}}{q_{\bullet \bullet}} \log \left( \frac{q_{ij} / q_{\bullet \bullet}}{p_{ij} / p_{\bullet \bullet}} \right)$ is minimum.
	\end{itemize}
	\end{frame}

@@

		In our context, it means that if we define an \imp{origin-destination affinity matrix} $\mathbf{S} = (s_{st})$ with 
			$$
				\mathbf{S} = \left( \begin{array}{ccccc}
				0 & 1 & 1 & \cdots & 1 \\
				0 & 0 & 1 & \cdots & 1 \\
				\vdots & \vdots & \ddots & \ddots & \vdots \\
				0 & 0 & \cdots & 0 & 1 \\
				0 & 0 & \cdots & 0 & 0 
				\end{array} \right)
			$$
		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e we find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verify:
		\begin{enumerate}
			\item $n_{s\bullet} = \rho^\text{in}_s$,
			\item $n_{\bullet t} = \rho^\text{out}_t$,
			\item $K(\mathbf{N}| \mathbf{S})$ is minimum.
		\end{enumerate}
		\small (a small number $\epsilon$ has to be added on null components).
	\end{frame}

@@

	\begin{frame}{Solution with iterative proportional fitting}
		By \imp{decreasing (resp. increasing) $s_{st}$}, we \imp{reduce (resp. expand)} the resulting number of passengers going from $s$ to $t$ obtained with IPF. \\
		\vspace{0.4cm}
		Thus, this approach is more \imp{flexible}, because we could give a \imp{specific affinity matrix $\mathbf{S} = (s_{st})$}, based on other data (additional assumptions).\\
		\vspace{0.4cm}
		The relationship $n_{ij} = a_s b_t s_{st}$ can be seen as a \imp{gravity model} (still to investigate).
	\end{frame}

@@

		\begin{center}
			\includegraphics[width=0.5\textwidth]{img/edge_type2.png}
		\end{center}

		We cannot use a Markov chain modeling in this case, but the \imp{iterative proportional fitting} approach is still promising. However, there are \imp{additional difficulties}.

@@

		\begin{itemize}
			\item $\rho^\text{in}_i = \sigma^\text{in}_i + x^\text{B}_{\bullet i}$,
			\item $\rho^\text{out}_i = \sigma^\text{out}_i + x^\text{B}_{i \bullet}$,
		\end{itemize}

		where $x^\text{B}_{\bullet i}$ is the \imp{transfer flow entering the node $i$} and $x^\text{B}_{i \bullet}$ the \imp{transfer flow leaving the node $i$}. \\
	\end{frame}

@@

	\begin{frame}{The flow in/out lines/network}
		The flow entering/leaving the lines, $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$, are \imp{known}. \\
		\vspace{0.4cm}
		If we know \imp{transfer flow on edges}, i.e. $\mathbf{X}_\text{B} = (x^\text{B}_{ij})$, we can \imp{compute the flow entering/leaving the network}, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{in}$. \\
		\vspace{0.4cm}
		The flow entering/leaving the lines, $\rho^\text{in}_i$ and $\rho^\text{out}_i$, can also acts as \imp{constraints on the in/out transfer flow}, $x^\text{B}_{\bullet i}$ and $x^\text{B}_{i \bullet}$.\\
		\vspace{0.4cm}
		When there are no transfers on $i$, we have $\rho^\text{in}_i = \sigma^\text{in}_i$ and  $\rho^\text{out}_i = \sigma^\text{out}_i$. \\
	\end{frame}

@@

		We now begin to make an outline for an \imp{iterative Algorithm}. There will be 4 steps at each iteration: \\
		\footnotesize
		\begin{align}
			\left. \begin{array}{r} 
				{\color{cyan} \text{OD affinity matrix } \mathbf{S}} \\
				{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
				{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
			\end{array} \right\}  &\overset{IPF}{\longrightarrow} \text{OD matrix } \mathbf{N} \\
			\begin{array}{r} ^
			\text{OD matrix } \mathbf{N}
			\end{array} &\overset{SP}{\longrightarrow} \text{Edge transfer flow } \mathbf{X}^\text{B} \\
			\left. \begin{array}{r} 
			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
			\text{Flow in lines } \bm{\rho}_\text{in} \\
			\text{Flow out lines } \bm{\rho}_\text{out}
			\end{array} \right\}  &\overset{\text{Constraints}}{\longrightarrow} 
			\left\{ \begin{array}{l}
			\text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
			{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
			{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
			\end{array} \right. \\
			\left. \begin{array}{r} 
			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
			 \text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B}
			\end{array} \right\}  &\overset{\text{Affinity update}}{\longrightarrow} {\color{cyan} \text{OD affinity matrix } \mathbf{S}}
		\end{align}	
	\end{frame}

@@

		It now possible to obtain origin-destination matrix $\mathbf{N}$ with \imp{iterative proportional fitting}.

@@

		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{corrected transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$. There are multiple choices:

		\begin{enumerate}
			\item \imp{Constraint thresholds} could be reachable.
			\item We can set a \imp{percentage limit} for the transfer flow among the flow in/out of the lines.
			\item We can set a \imp{soft limit} to the transfer flow, with, e.g. an exponential law.
		\end{enumerate}

		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with $\sigma^\text{in}_i = \rho^\text{in}_i - \widetilde{x}^\text{B}_{\bullet i}$ and $\sigma^\text{out}_i = \rho^\text{out}_i - \widetilde{x}^\text{B}_{i \bullet}$.

@@

		What about the \imp{excess flow on transfer edges}, $x^\text{B}_{ij} - \widetilde{x}^\text{B}_{ij}$ ? \\
		\vspace{0.4cm}
		Note that we only affected \imp{margins distribution}, but we need also need a way to reduce \imp{particular dependencies} between some $s$ and $t$. \\
		\vspace{0.4cm}
		Having a \imp{shortest-path structure}, we know which couples $s, t$ are responsible for the excess flow on edge $(i, j)$.
		\begin{center}
			\includegraphics[width=0.6\textwidth]{img/excess_flow.png}
		\end{center}
	\end{frame}

@@

		On each transfer edge, we can compute the \imp{proportion of allowed flow}:
		$$
			p^\text{allowed}_{ij} = \frac{\widetilde{x}_{ij}^\text{B}}{x_{ij}^\text{B}}
		$$
		Each transfer edge will then "send a signal" to all couples $(s,t)$ using this edge, in order for them to \imp{reduce their affinities $s_{st}$}. \\
		\vspace{0.4cm}
		Each couple $(s,t)$, will recieve a \imp{list} of proportion of allowed flow, from all transfer edges on its shortest-path. The \imp{update factor for the affinity $s_{st}$} is constructed by using the minimum allowed flow received.
		$$
			s^\text{new}_{st} = s^\text{old}_{st} \cdot (1 - \min\{p^\text{allowed}_{i_1, j_1}, \ldots, p^\text{allowed}_{i_k, j_k} \})
		$$
	\end{frame}

	\section{Results}

	\begin{frame}[allowframebreaks]{References}

		\bibliography{demo}
		\bibliographystyle{abbrv}

>>>>

\title{Estimation of flow trajectories in a multiple lines network}
\subtitle{Experiments with \textit{transports publics de la région lausannoise} (tl) data}
\date{}
\author{Guillaume Guex \\ Romain Loup \\ François Bavaud}
\institute{University of Lausanne}

@@

	\begin{frame}{Problematic}

@@

	\begin{frame}

@@

	\begin{frame}{Formal problem definition}
		Let a line (in one direction), which have $n$ stops, indexed by line order. Let $\bm{\rho}_\text{in} = (\rho^\text{in}_s)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_t)$ be two vectors representing, respectively, the \imp{passengers entering and leaving lines at each stop}. \\
		\vspace{0.4cm}
		We search a $(n \times n)$ \imp{origin-destination matrix $\mathbf{N} = (n_{st})$} where components represents
		$$
			{\color{cyan} n_{st} = \text{"the number of passengers entering line at $s$ and leaving at $t$"}}.
		$$
		These components must verify
		\begin{enumerate}
			\item $n_{st} \geq 0$,
			\item $n_{s\bullet} = \rho^\text{in}_s$,
			\item $n_{\bullet t} = \rho^\text{out}_t$.
		\end{enumerate}
		\small ($\bullet$ indicates a sum on the replaced index)
	\end{frame}

@@

	The same solution can be obtained with the \imp{iterative proportional fitting (IPF)} algorithm [Bishop et al., 1975]. Let 
	\begin{enumerate}
		\item $\mathbf{P} = (p_{ij})$ a $(n \times m)$ matrix, 
		\item $\mathbf{u} = (u_i)$ a n-length vector, and 
		\item $\mathbf{v} = (v_i)$ a m-length vector,
	\end{enumerate}
	all of them with strictly positive components. We can find two vectors $\mathbf{a} = (a_i)$ and $\mathbf{b} = (b_i)$ such that the matrix $\mathbf{Q} = (q_{ij})$, defined with
	$$
		q_{ij} = a_i b_j p_{ij},
	$$
	verifies
	\begin{itemize}
		\item $q_{i\bullet} = u_i$,
		\item $q_{\bullet j} = v_j$,
		\item $K(\mathbf{Q}| \mathbf{P}) := \sum_{ij} \frac{q_{ij}}{q_{\bullet \bullet}} \log \left( \frac{q_{ij} / q_{\bullet \bullet}}{p_{ij} / p_{\bullet \bullet}} \right)$ is minimum.
	\end{itemize}
	\end{frame}

@@

		In our context, it means that if we define an \imp{origin-destination affinity matrix $\mathbf{S} = (s_{st})$} with 
			$$
				\mathbf{S} = \left( \begin{array}{ccccc}
				0 & 1 & 1 & \cdots & 1 \\
				0 & 0 & 1 & \cdots & 1 \\
				\vdots & \vdots & \ddots & \ddots & \vdots \\
				0 & 0 & \cdots & 0 & 1 \\
				0 & 0 & \cdots & 0 & 0 
				\end{array} \right)
			$$
		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e., we can find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verifies
		\begin{enumerate}
			\item $n_{s\bullet} = \rho^\text{in}_s$,
			\item $n_{\bullet t} = \rho^\text{out}_t$,
			\item $K(\mathbf{N}| \mathbf{S})$ is minimum.
		\end{enumerate}
		\small (a small number $\epsilon$ has to be added on null components).
	\end{frame}

@@

	\begin{frame}{Solution with iterative proportional fitting}
		By \imp{decreasing (resp. increasing) $s_{st}$}, we \imp{reduce (resp. expand)} the resulting number of passengers going from $s$ to $t$ obtained with IPF. \\
		\vspace{0.4cm}
		Thus, this approach is more \imp{flexible}, because we could give a \imp{specific affinity matrix $\mathbf{S} = (s_{st})$}, based on other data (additional assumptions).\\
		\vspace{0.4cm}
		The relationship $n_{st} = a_s b_t s_{st}$ can be seen as a \imp{gravity model} (still to investigate).
	\end{frame}

@@

		\begin{center}
			\includegraphics[width=0.45\textwidth]{img/edge_type2.png}
		\end{center}

		It is not possible to use a Markov chain modeling in this case, but the \imp{iterative proportional fitting} approach is still valid. However, there are \imp{additional difficulties}.

@@

		\begin{align*}
			\rho^\text{in}_i &= \sigma^\text{in}_i + x^\text{B}_{\bullet i}, \\
			\rho^\text{out}_i &= \sigma^\text{out}_i + x^\text{B}_{i \bullet},
		\end{align*}

		where 
		\begin{itemize}
			\item $x^\text{B}_{\bullet i}$ is the \imp{transfer flow entering the node $i$} and
			\item $x^\text{B}_{i \bullet}$ is the \imp{transfer flow leaving the node $i$}.
		\end{itemize}
	\end{frame}

@@

	\begin{frame}{The flow in/out lines/network}
		The flow entering/leaving the lines, $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$, are \imp{known}. \\
		\vspace{0.4cm}
		If we know \imp{transfer flow on edges}, i.e. $\mathbf{X}_\text{B} = (x^\text{B}_{ij})$, we can \imp{compute the flow entering/leaving the network}, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{in}$. \\
		\vspace{0.4cm}
		The flow entering/leaving the lines, $\rho^\text{in}_i$ and $\rho^\text{out}_i$, acts as \imp{constraints on the in/out transfer flow}, $x^\text{B}_{\bullet i}$ and $x^\text{B}_{i \bullet}$.\\
		\vspace{0.4cm}
		When there are no transfers on $i$, we have $\rho^\text{in}_i = \sigma^\text{in}_i$ and  $\rho^\text{out}_i = \sigma^\text{out}_i$. \\
	\end{frame}

@@

		Let us make an outline for the \imp{iterative Algorithm}. There is \imp{4 steps} at each iteration: \\
		\footnotesize
		\begin{align}
			\left. \begin{array}{r} 
				{\color{cyan} \text{OD affinity matrix } \mathbf{S}} \\
				{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
				{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
			\end{array} \right\}  &\overset{IPF}{\longrightarrow} \text{OD matrix } \mathbf{N} \\
			\begin{array}{r}
			\text{OD matrix } \mathbf{N}
			\end{array} &\overset{SP}{\longrightarrow} \text{Transfer flow } \mathbf{X}^\text{B} \\
			\left. \begin{array}{r} 
			\text{Transfer flow } \mathbf{X}^\text{B} \\
			\text{Flow in lines } \bm{\rho}_\text{in} \\
			\text{Flow out lines } \bm{\rho}_\text{out}
			\end{array} \right\}  &\overset{\text{Constraints}}{\longrightarrow} 
			\left\{ \begin{array}{l}
			\text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
			{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
			{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
			\end{array} \right. \\
			\left. \begin{array}{r} 
			\text{Transfer flow } \mathbf{X}^\text{B} \\
			 \text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B}
			\end{array} \right\}  &\overset{\text{Affinity update}}{\longrightarrow} {\color{cyan} \text{OD affinity matrix } \mathbf{S}}
		\end{align}	
	\end{frame}

@@

		It now possible to obtain a first origin-destination matrix $\mathbf{N}$ with \imp{iterative proportional fitting}.

@@

	\begin{frame}{Step 3: allowed transfer flow}

		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{allowed transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$. There are multiple choices:

		\begin{enumerate}
			\item \imp{Constraint thresholds} could be reachable.
			\item We can set a \imp{percentage limit} for the transfer flow among the flow in/out of the lines.
			\item We can set a \imp{soft limit} to the transfer flow, with, e.g., an exponential law.
		\end{enumerate}

		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with 
		\begin{align*}
			\sigma^\text{in}_i &= \rho^\text{in}_i - \widetilde{x}^\text{B}_{\bullet i}, \\	\sigma^\text{out}_i &= \rho^\text{out}_i - \widetilde{x}^\text{B}_{i \bullet}.
		\end{align*}
	\end{frame}

		What about the \imp{excess flow on transfer edges}, i.e., $x^\text{B}_{ij} - \widetilde{x}^\text{B}_{ij}$ ? \\
		\vspace{0.4cm}
		Note that we only updated \imp{margins distribution}, but we need also need a way to reduce \imp{dependencies} between particular $s$ and $t$. \\
		\vspace{0.4cm}
		Having flow following \imp{shortest-paths}, we know which couples $s, t$ are "responsible" for the excess flow on edge $(i, j)$.
		\begin{center}
			\includegraphics[width=0.6\textwidth]{img/excess_flow.png}
		\end{center}
	\end{frame}

		On each transfer edge, we can compute the \imp{proportion of allowed flow}:
		$$
			p^\text{allowed}_{ij} = \frac{\widetilde{x}_{ij}^\text{B}}{x_{ij}^\text{B}}
		$$
		Each transfer edge will then "send a signal" to all couples $s,t$ using this edge on their shortest-paths, in order to \imp{reduce their affinities $s_{st}$}. \\
		\vspace{0.4cm}
		Each couple $s,t$, will recieve a \imp{list of proportion of allowed flow}, from all transfer edges on its shortest-path. The \imp{update factor for the affinity $s_{st}$} is constructed by using the minimum allowed flow received.
		$$
			s^\text{new}_{st} = \min\{p^\text{allowed}_{i_1, j_1}, \ldots, p^\text{allowed}_{i_k, j_k} \} \cdot s^\text{old}_{st}
		$$
	\end{frame}

	\begin{frame}{Step 4: Affinity update}
		\begin{center}
			\includegraphics[width=0.7\textwidth]{img/allowed_flow.png} \\
			In this example, $s^\text{new}_{st} = 0.3 \cdot s^\text{old}_{st}$
		\end{center}
	\end{frame}

	\section{Results (demo)}

	\section{Conclusion}

	\begin{frame}{Conclusion}

		As it is still a work in progress, the conclusion will take the form of a to-do list:
		\begin{itemize}
			\item \imp{Optimizing} speed/memory (almost done).
			\item Running the algorithm on \imp{all lines}.
			\item Using carefully crafted \imp{OD affinities}. 
			\item Finding \imp{applications}.
			\item Finding a \imp{proof of convergence}. 
			\item Finding if the solution is \imp{unique} (doubt it).
			\item Making links with \imp{gravity model}, \imp{transportation problem}, ...
			\item \imp{Adapting} the algorithm to closely related problems.
		\end{itemize}

@@

	\begin{frame}

		\begin{center}
			\huge Thank you for your attention ! \\
			Questions ? \\
			\vspace{1cm}
			\small
			(for real this time)
		\end{center}

@@
